Measuring contrast processing in the visual system using SSVEP

### 2024-05-08 review.qmd
All

This approach is primarily useful in situations that require the BCI to distinguish from among a small set of possibilties: for example, in early work, visual stimuli representing 'Left' and 'Right' commands in a flight simulator were distinguished robustly [@Middendorf2000]. Although SSVEP-based BCI interfaces typically do use contrast flicker, work in this field has largely focused on optimising the stimuli or decoders to increase decoding performance (e.g. ... ) and reducing visual fatigue associated with the long-term presentation of arrays of high-contrast flicker (e.g. .....) rather than studying visual contrast processing \textit{per se}. We therefore note in passing that using SSVEPs to improve our understanding of early contrast processing may yield benefits to this related field.

# Future directions
The SSVEP is a powerful tool for studying contrast processing. It provides a high SNR readout of neuronal activity that is unambigously linked to the input. It is sensitive to both the amplitude and phase of the input and when combined with source imaging, it can be extracted from different cortical regions allowing researchers to track contrast processing computations across the visual pathway. Because of the high SNR, it can be measured in subjects where long recording durations are impractical (for example, infants or patients with neurological disorders) and at frequencies high enough to be effectively invisible [REF - Some high frequency work... - Romanian group]. The SSVEP is also able to 'fingerprint' modulators of the inputs through the harmonic and intermodulation terms they generate in the output. In principle, each nonlinearity in the visual pathway can be identified by its contribution to the frequency spectrum at different recording locations. This, it turn, allows researchers to study how contrast processing depends on spatial and temporal context and as well as changes in task, cognitive and behavioural state and arousal.

Although visual neuroscience is a relatively old field, there are still outstanding questions relating to contrast processing that could be addressed by SSVEP methods. For example, it is still not completely clear how contrast signals are computed in the human retina: although we have more than a century of electrophysiological data from animals, and the broad structure of cone inputs to retinal ganglion cells is understood [@LiFieldNeuron2014] we are still discovering new aspects of retinal processing that could influence the 'coining' of the visual system's currency [@gollischMeister2010, @WangPuthussery2023Nature, @uprety2022melanopsin]. The ability to frequency tag both inputs (for example, cone-directed luminance contrast) and modulators (for example, stimuli that selectively drive the intrinsically-photosensitive retinal ganglion cells) is a powerful tool to explore this first stage of image generation.

@@

Finally...

>>>>

This approach is primarily useful in situations that require the BCI to distinguish from among a small set of possibilties: for example, in early work, visual stimuli representing 'Left' and 'Right' commands in a flight simulator were distinguished robustly [@Middendorf2000]. Although SSVEP-based BCI interfaces typically do use contrast flicker, work in this field has largely focused on optimising the stimuli or decoders to increase decoding performance (e.g. ... ) and reducing visual fatigue associated with the long-term presentation of arrays of high-contrast flicker (@Diez2024AssessmentMedBiolComp) rather than studying visual contrast processing \textit{per se}. We therefore note in passing that using SSVEPs to improve our understanding of early contrast processing may yield benefits to this related field.

# Future directions
The SSVEP is a powerful tool for studying contrast processing. It provides a high SNR readout of neuronal activity that is unambigously linked to the input. It is sensitive to both the amplitude and phase of the input and when combined with source imaging, it can be extracted from different cortical regions allowing researchers to track contrast processing computations across the visual pathway. Because of the high SNR, it can be measured in subjects where long recording durations are impractical (for example, infants or patients with neurological disorders) and at frequencies high enough to be effectively invisible [@Hermann2001ExpBrRes, @MinarikNeuroimage2023]. The SSVEP is also able to 'fingerprint' modulators of the inputs through the harmonic and intermodulation terms they generate in the output. In principle, each nonlinearity in the visual pathway can be identified by its contribution to the frequency spectrum at different recording locations[@Regan1988]. This, it turn, allows researchers to study how contrast processing depends on spatial and temporal context and as well as changes in task, cognitive and behavioural state and arousal.

Although visual neuroscience is a relatively old field, there are still outstanding questions relating to contrast processing that could be addressed by SSVEP methods. For example, it is still not completely clear how contrast signals are computed in the human retina. Although we have more than a century of electrophysiological data from animals, and the broad structure of cone inputs to retinal ganglion cells is understood [@LiFieldNeuron2014], we are still discovering new aspects of retinal processing that could influence the 'coining' of the visual system's currency [@gollischMeister2010, @WangPuthussery2023Nature, @uprety2022melanopsin]. The ability to frequency tag both inputs (for example, cone-directed luminance contrast) and modulators (for example, stimuli that selectively drive the intrinsically-photosensitive retinal ganglion cells) is a powerful tool to explore this first stage of image generation.

@@

Finally, SSVEPs continue to provide a way of studying contrast in the cortex. Here, we are often interested in how different visual parameters interact. For example, how are signals from different eyes combined to generate both scalar contrast values and also binocular depth cues? How does contrast combination depend on the low level properties of the individual inputs such as their retinotopic location, cone contrast, spatiotemporal frequency, eye of origin and orientation? Colour vision scientists are particularly interested in how chromatic signals originating in a small number of cone-opponent retinal pathways are transformed into a perceptual colour space where the 'unique hues' appear to be only weakly-related to the early retinal outputs and how these computations are conditioned by the spatial and temporal properties of the scene [@LiConeOppNatComm22, @ConwayUniqueHues,@KanekoCerebCort2020, @GegenfurtnerNatRev2003, @SolomonLennie2007NatRev, @WandellColorPNAS1993, ]. All of these questions can be addressed by using SSVEP and frequency tagging to examine the computations that combine and transform the inputs at different cortical stages (@WattsBiorxiv2023https://www.biorxiv.org/content/10.1101/2023.12.10.570997v2, @BakerOptimalSignal2017, @ChenGegenfurtnerJOV2021, @KatyalDSciRep2018, @Busse2009). The original promise of the SSVEP approach was that the  entire complex-valued frequency spectrum recorded at each lcoation provided detailed information about the processing nonlinearities up to and including that point. In principle, this would allow researchers to uniquely fit the parameters of their computations models of visual processing [@Regan1988].  To date, the complexity of the neuronal computations at even the earliest stage of visual processing have hampered this effort - researchers typically restrict their analyses to the amplitudes of single, low-order frequency components (for example, the inputs frequencies or simple sums and differences of those frequencies). As our understanding of the early visual system improves, it is becoming possible to generate more realistic parameterised forward models of signal generation [@CharikerJOV2020, @Winawer, @Baker, @Tsai2012, OTHERS from Winawer]. Feeding frequency-tagged inputs into these types of model allows us to generate synthetic SSVEP responses that can be compared with those measured in human subject. In principle, we are therefore able to use the SSVEP to fit the parameters of early visual processing. This approach may also allow us to develop more sensitive tests for the changes in early visual processing that accompany a wide range of neurological diseases and disorders.
